---
title: "RlabFinal"
author: "Shilpi Jaiswal"
date: "March 16, 2016"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

library("dplyr")  #for dplyr
library("RSQLite") #for sqllite
library("nycflights13") #for data
library(ggplot2)

my.db <- nycflights13_sqlite()

weather.df  <- weather %>% collect()
planes.df  <- planes %>% collect()
flights.df <- tbl(my.db, "flights") %>% 
                 collect() %>%
                 
                  mutate(
                       
                     cancelled = ifelse(is.na(arr_time), 1, 0), # 151 cancellations
                     delayed =  ifelse(dep_delay > 0, 1, 0),
                     date = paste(year, month, day, sep = "-"),
                     date = as.Date(date, "%Y-%m-%d")
                  )

```


```{r, echo=FALSE, warning=FALSE}

weather.df <- weather.df %>%
             mutate(
                  
                  date = paste(year, month, day, sep = "-"),
                  date = as.Date(date, "%Y-%m-%d")
   
             )
```


```{r, echo=F, message=FALSE}

flt.plane <- flights.df %>% 
              left_join(planes.df, by = "tailnum")

flt.plane.w <- flt.plane %>% inner_join(weather.df, by = "date")  %>% 
      select(-year.x, -month.x, -day.x, -precip, -hour.x, -hour.y, -manufacturer,-type, -speed)

```

#### Part A
```{r,echo=FALSE, warning=FALSE}
cancel.w <- glm(cancelled ~ dep_delay+temp+wind_speed+visib, family=binomial(link="probit"), data = flt.plane.w)
```

```{r, eval=TRUE}
summary(cancel.w)
```
This probit regression shows that for cancelled flights, delay in departure, temperature and visibility are the most significant factors. We can also see that the delay in departure also significantly increases the probability of cancellation of a flight.

***
##### Effect of weather on cancellation
```{r, echo=FALSE, fig.height = 3., fig.width=6, message=FALSE, fig.cap = "Density plot: Dew Point, on Cancelled flights", warning=FALSE}

ggplot(flt.plane.w, aes(x = dewp, y = ..density..)) + 
  geom_histogram(binwidth = 1) +
  facet_grid(cancelled ~ .)

```

##### Effect of weather on delay 
```{r, echo=FALSE, message=FALSE, fig.cap = "Density plot: Dew Point, on Delayed flights"}

 ggplot(flt.plane.w, aes(x = dewp, y = ..density..)) + 
  geom_histogram(binwidth = 1) +
  facet_grid(delayed ~ .)

```


```{r, echo=FALSE, message=FALSE, fig.cap = "Plot of median departure delay by visibility"}
# flt.plane.w; distinct(select(flt.plane.w, visib))

qplot(dep_delay, data=flt.plane.w, geom="density", fill=visib, alpha=I(.5),
      main="Distribution of Departure Delays, by visibility", xlab="Departure Delay",
      ylab="Density")
```

***
#### Part B

##### Analysis on aspect of time, on delay and cancellation
```{r, echo=F}
time_analysis <- flt.plane.w %>% group_by(cancelled, year.y, month.y, day.y) %>%
  mutate(
    
    delay.median = median(dep_delay, na.rm = T),
    count = n()
  ) %>% 
  select(year.y, month.y,day.y, date, delay.median) %>%
  arrange(cancelled, year.y, month.y, day.y)

```

```{r, echo=FALSE, message=FALSE, fig.cap = "Plot of median departure delay"}
plot2a <- ggplot(time_analysis, aes( x = date, y = delay.median), main="Median departure delay by day", xlab="Date", ylab="Median  Departure Delay") + 
  geom_line(aes(color = cancelled), size = 1.05)

plot2a
```
From the above plot we can see that most of the flights cancelled have higher delays, which might mean the flights were finally cancelled once the flight couldnt be started after a long delay.

```{r, echo=FALSE, message=FALSE, fig.cap = "Plot of median departure delay"}
plot2b <- ggplot(time_analysis, aes( x = date, y = delay.median), main="Median departure delay by day", xlab="Date", ylab="Median  Departure Delay") + geom_line(aes(color = delay.median), size = 1.05)

plot2b
```
Here I have used colour by median delay for a given day, we can see that the longer departure delays are mostly 
between the month of April and July, unlike the cancelled flights are more frequent between months of May to August.
We can see that in those months the probability of a delayed flight being cancelled is higher.


#### Part C  Relationship between delay and destination.
```{r, echo=FALSE, message=FALSE, fig.cap = "Plot of departure delay  by airport destination"}
qplot(dep_delay, data=flt.plane.w, geom = "density", fill=dest, alpha=I(.5),
      main="Distribution of Departure Delays, by airport destination", xlab="Departure Delay",
      ylab="Density")

```

This plot shows the delayed flights by destination.


#### Part C  Relationship between delay and manufacturing year of the planes.
```{r, echo=FALSE, message=FALSE, fig.cap = "Plot of departure delay  by planes manufacturing years"}
qplot(dep_delay, data=flt.plane.w, geom = "histogram", fill=year.y, alpha=I(.5), 
      main="Departure Delays by planes manufacturing years", xlab="Departure Delay", ylab="Density")

```

This plot shows delay by manufacturing years of the planes.